Sorting by Reversals

### 2020-06-30 slides.tex
improvements

\begin{definition}[reversal distance problem]
Given two permutations $\sigma, \tau \in S_n$ find a sequence of reversals $\rho_1, \dots, \rho_d \in S_n$ such that
\begin{align*}
    \sigma \circ \rho_1 \circ \cdots \circ \rho_d = \tau
\end{align*}
and $d$ is minimal. \pause\par\padding
$d$ is called \b{reversal distance} between $\sigma$ and $\tau$. \cite{Kececioglu1995}
\end{definition}

@@

\begin{definition}[MIN-SBR]
Let $\pi = \tau^{-1} \circ \sigma \in S_n$. \par\pause
\textit{Sorting by Reversals} is the problem of finding a sequence of reversals $\rho_1, \dots, \rho_d \in S_n$ such that
\begin{align*}
    \pi \circ \rho_1 \circ \cdots \circ \rho_d = id
\end{align*}
and $d$ is minimal. \cite{Kececioglu1995}
\end{definition}

@@

\begin{exampleblock}{Example}
Let $\pi = (0\ 4\ 2\ 1\ 3\ 5) \in S_4$.\pause
\begin{align*}
    \pi \circ \rho(1,3) &= (0\ \r{1\ 2\ 4}\ 3\ 5) \\ \pause
    \pi \circ \rho(1,3) \circ \rho(3,4) &= (0\ 1\ 2\ \r{3\ 4}\ 5) = id
\end{align*}\pause
$\implies d(\pi) \leq 2$.
\end{exampleblock}\pause\padding

\begin{algorithm}[ratchet algorithm, \citeauthor*{Watterson19821}]
For each element at position $i$, perform $\rho(i, \pi_i^{-1})$ if $i \neq \pi_i$.
\end{algorithm}\pause

\begin{corollary}[tight upper bound, \citeauthor*{Kececioglu1995}]
$d(\pi) \leq n - 1$ for all $\pi \in S_n$.
\end{corollary}

@@

Observation: $b(\pi) = 0$ iff $\pi = id$ and any reversal can at most eliminate two breakpoints. \pause\spadding

@@

\begin{definition}[breakpoint graph, \citeauthor*{Bafna1996}]
Let $G(\pi) = (V, E)$ with
\begin{itemize}
    \item vertices $V = [0, n+1]$ representing the elements of $\pi$\pause; and
    \item edges $E = \r R \cup \b B$ with
    \begin{itemize}
        \item a \r{red} edge for every breakpoint in $\pi$; and
        \item a \b{blue} edge for every adjacency in $\pi$.
    \end{itemize}
\end{itemize}
\end{definition}

@@

\begin{exampleblock}{Example}
Let $\pi = (0\ \r\mid\ 4\ \r\mid\ 2\ 1\ \r\mid\ 3\ \r\mid\ 5) \in S_4$.

@@

@@

A reversal \b{acts on} two red edges of $G\pi)$ if those two edges represent the breakpoints that are split apart by the reversal. \pause\spadding

@@

\begin{definition}[reversal graph of $G(\pi)$, \citeauthor*{Christie1998}]
Given an alternating cycle decomposition $C$ of $G(\pi)$, \par\pause
define $R(C)$ with
\begin{itemize}
    \item an isolated \b{blue} vertex for each adjacency in $\pi$\pause;
    \item $m$ vertices for each $m$-cycle in $C$ each representing the reversal $\rho(u)$ acting on the two red edges connected by a blue edge\pause;
    \begin{itemize}
        \item a vertex is \r{red} if the represented reversal is a $1$- or $2$-reversal
        \item a vertex is \b{blue} otherwise
    \end{itemize}\pause
    \item connect two vertices with an edge if their corresponding blue edges \textit{interleave} (i.e. overlap).
\end{itemize}
\end{definition}\pause

@@

\begin{definition}
Denote by $C_u$ the alternating cycle decomposition of $G(\pi \circ \rho_u)$ that is obtained from $C$ by removing one 2-cycle or shortening one of its cycles by one red edge.
\end{definition}\pause

>>>>

\begin{definition}[reversal distance problem]
Given two permutations $\sigma, \tau \in S_n$ find a sequence of reversals $\rho_1, \dots, \rho_d \in S_n$ such that
\begin{align*}
    \sigma \circ \rho_1 \circ \cdots \circ \rho_d = \tau
\end{align*}
and $d$ is minimal. \pause\par\padding
$d$ is called \b{reversal distance} between $\sigma$ and $\tau$.
\end{definition}

@@

\begin{definition}[MIN-SBR]
Let $\pi = \tau^{-1} \circ \sigma \in S_n$. \par\pause
\textit{Sorting by Reversals} is the problem of finding a sequence of reversals $\rho_1, \dots, \rho_d \in S_n$ such that
\begin{align*}
    \pi \circ \rho_1 \circ \cdots \circ \rho_d = id
\end{align*}
and $d$ is minimal.
\end{definition}

@@

\begin{exampleblock}{Example}
Let $\pi = (0\ 4\ 2\ 1\ 3\ 5) \in S_4$.\pause
\begin{align*}
    \pi \circ \rho(1,3) &= (0\ \r{1\ 2\ 4}\ 3\ 5) \\ \pause
    \pi \circ \rho(1,3) \circ \rho(3,4) &= (0\ 1\ 2\ \r{3\ 4}\ 5) = id
\end{align*}\pause
$\implies d(\pi) \leq 2$.
\end{exampleblock}

% \begin{algorithm}[ratchet algorithm, \citeauthor*{Watterson19821}]
% For each element at position $i$, perform $\rho(i, \pi_i^{-1})$ if $i \neq \pi_i$.
% \end{algorithm}\pause

% \begin{corollary}[tight upper bound, \citeauthor*{Kececioglu1995}]
% $d(\pi) \leq n - 1$ for all $\pi \in S_n$.
% \end{corollary}

@@

Observation: $b(\pi) = 0$ iff $\pi = id$ \pause and any reversal can at most eliminate two breakpoints. \pause\spadding

@@

\begin{definition}[breakpoint graph, \citeauthor*{Bafna1996}]
Let $G(\pi) = (V, E)$ with
\begin{itemize}
    \item vertices $V = [0, n+1]$ representing the elements of $\pi$\pause; and
    \item edges $E = \r R \cup \b B$ with
    \begin{itemize}
        \item a \r{red} edge for every breakpoint in $\pi$; and
        \item a \b{blue} edge for every missing adjacency in $\pi$.
    \end{itemize}
\end{itemize}
\end{definition}

@@

\begin{exampleblock}{Example}
Let $\pi = (0\ \r\mid\ 4\ \r\mid\ 2\ 1\ \r\mid\ 3\ \r\mid\ 5) \in S_4$. \pause Then $G(\pi)$ is

@@

\begin{center}
\begin{tikzpicture}[scale=0.6, every node/.style={scale=0.6}]
	\begin{pgfonlayer}{nodelayer}
		\node [style=node] (0) at (-3, 0) {0};
		\node [style=node] (3) at (-1.5, 0) {4};
		\node [style=node] (4) at (0, 0) {2};
		\node [style=node] (5) at (1.5, 0) {1};
		\node [style=node] (8) at (3, 0) {3};
		\node [style=node] (9) at (4.5, 0) {5};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw [style=red edge] (3) to (4);
		\draw [style=red edge] (8) to (9);
		\draw [style=red edge] (0) to (3);
		\draw [style=red edge] (5) to (8);
		\draw [style=blue edge, bend left] (4) to (8);
		\draw [style=blue edge, bend left] (3) to (9);
		\draw [style=blue edge, bend right] (0) to (5);
		\draw [style=blue edge, bend right] (3) to (8);
	\end{pgfonlayer}
\end{tikzpicture}
\end{center}

@@

A reversal \b{acts on} two red edges of $G(\pi)$ if those two edges represent the breakpoints that are split apart by the reversal. \pause\spadding

@@

\begin{definition}[reversal graph, \citeauthor*{Christie1998}]
Given an alternating cycle decomposition $C$ of $G(\pi)$, \par\pause
define $R(C)$ with
\begin{itemize}
    \item an isolated \b{blue} vertex for each adjacency in $\pi$\pause;
    \item $m$ vertices for each $m$-cycle in $C$ each representing the reversal $\rho(u)$ acting on the two red edges connected by a blue edge\pause;
    \begin{itemize}
        \item a vertex is \r{red} if the represented reversal is a $1$- or $2$-reversal
        \item a vertex is \b{blue} otherwise
    \end{itemize}\pause
    \item connect two vertices with an edge if their corresponding blue edges \textit{interleave} (i.e. overlap).
\end{itemize}
\end{definition}\pause

@@

\begin{definition}
Denote by $C_u$ the alternating cycle decomposition of $G(\pi \circ \rho(u))$ that is obtained from $C$ by removing one 2-cycle or shortening one of its cycles by one red edge.
\end{definition}\pause
